Old or Heavy?\\Decaying Gracefully with Age/Weight Shapes.

### 2019-03-02 papers/CADE19/paper.tex
proof reading

\documentclass{llncs}
\usepackage{graphicx}
\usepackage[caption=false]{subfig}
\usepackage[dvipsnames]{xcolor}
\usepackage{amssymb}
\usepackage{amsmath}
\usepackage{todonotes}
%\usepackage{multicol}
%\newcommand{\todo}[1]{}

@@

\begin{document}
\maketitle
\begin{abstract}
Modern saturation-based theorem provers are based on the given-clause algorithm, which iteratively selects new clauses to process. This clause selection has a large impact on the performance of proof search and has been the subject of much folklore.  
The standard approach is to alternate between selecting the \emph{oldest} clause and the \emph{lightest} clause with a fixed, but configurable \emph{age/weight ratio} (AWR).
We show that an optimal fixed value of this ratio can produce proofs significantly more quickly on a given problem, and further that varying AWR during proof search can improve upon a fixed ratio.
Based on these observations we develop several new modes for the Vampire prover  which vary AWR according to a ``shape'' during proof search.
The modes solve a number of new problems in the TPTP benchmark.
\end{abstract}

@@

Currently, the most successful theorem provers (such as Vampire~\cite{vampire}, E~\cite{E}, or SPASS~\cite{Spass}) for first-order logic are saturation-based, utilising the well-known \emph{given-clause algorithm}.
Simply, this algorithm \emph{saturates} a set of clauses by iteratively selecting a clause and performing all non-redundant inferences with it until all clauses have been selected or the empty clause (witnessing inconsistency) has been found.
Clearly, the order in which clauses are selected is key to the performance of the algorithm.
Over the past few decades a certain amount of folklore has built up around the best methods for clause selection and recent work by \sandm{} \cite{clause-selection-heuristics} systematically studied these.
Our work extends this study with new results and also introduces the concept of a \emph{variable} clause selection strategy (one that changes over time) which we instantiate with two simple patterns (or shapes) that prove to be pragmatically useful.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Background}
\label{sec:background}

This section introduces the relevant background for the rest of the paper.

%%%%%%
\paragraph{First-Order Logic and Weight.} Our setting is the standard first-order predicate logic with equality. A formal definition of this logic is not required for this paper but an important notion is that of the \emph{weight} of a clause. In first-order logic,  terms are built from function symbols and variables, literals are built from terms, and clauses are disjunctions of literals. The weight of a term is the number of function and variable symbols occurring in it and this notion is lifted to clauses.

\iffalse
Terms are of the form $f(t_1,\ldots,t_n), c$ or $x$ where
$f$ is a \emph{function symbol} of arity $n \geq 1$, $t_1,\ldots, t_n$ are terms, $c$ is a zero arity function symbol (i.e. a constant) and $x$ is a variable. 
%
The \emph{weight} of a term $t$ is defined as $w(t) =1$ if $t$ is a variable or a constant
and as $w(t) = 1+\sum_{i=1,\ldots,n} w(t_i)$ if $t$ is of the form
$f(t_1,\ldots,t_n)$. In other words, the weight or a term is the
number of symbols in it.
%
Atoms are of the form $p(t_1,\ldots,t_n), q$ or $t_1 \eql t_2$ where $p$ is a \emph{predicate symbol} of arity $n$, $t_1, \ldots, t_n$ are terms, $q$ is a zero arity predicate symbol and $\eql$ is the \emph{equality symbol}. 
%
The weight function naturally extends to atoms: $w(p(t_1,\ldots,t_n)) = 1+\sum_{i=1,\ldots,n} w(t_i)$, $w(t_1 \eql t_2) = w(t_1)+w(t_2)$,
and $w(q)=1$.
%
A literal is either an atom $A$, in which case we call it \emph{positive}, or a negation $\notl A$,
in which case we call it \emph{negative}.
%We write negated equalities as $t_1 \neql t_2$. 
%
The weight of a literal is the weight of the corresponding atom. 
% 
%We write $t[s]_p$ and $L[s]_p$ to denote that a term $s$ occurs in a term $t$ (in a literal $L$) at a position $p$.
%
A \emph{clause} is a disjunction of literals $L_1 \orl \ldots \orl L_n$ for $n \geq 0$.
We disregard the order of literals and treat a clause as a multiset. 
%
The weight of a clause is the sum of the weight of its literals. 
\fi

%%%%%%
\paragraph{Saturation-Based Proof Search.}

Saturation-based theorem provers \emph{saturate} a set of clauses $S$ with respect to an inference system $\mathbb{I}$ e.g. they compute a set of clauses $S'$ by applying rules in $\mathbb{I}$ to clauses in $S$ until no new clauses are generated. 
%
If the empty clause is generated then $S$ is unsatisfiable.
%
Calculi such as resolution and superposition have conditions that ensure \emph{completeness}, which means that a saturated set $S$ is satisfiable if it does not contain the empty clause as an element.
%
As first-order logic is only semi-decidable, it is not necessarily the case that $S$ has a finite saturation, and even if it does it may be unachievable in practice using the available resources. 
%
Therefore, much effort in saturation-based first-order theorem proving involves 
controlling proof search to make finding the empty clause more likely (within reasonable resource bounds). 
%
One important notion is that of \emph{redundancy}, being able to remove clauses from the search space that are not required. 
%
Another important notion are literal selections that place restrictions on the inferences that can be performed. 
%
Both notions come with additional requirements for completeness. Vampire often gives up these requirements for pragmatic reasons (incomplete strategies have been found to be more efficient than complete ones in certain cases) and in such cases the satisfiability of $S$ upon saturation is \emph{unknown}.

%%%%%%
\paragraph{The Given Clause Algorithm and AWR Clause Selection.}
To achieve saturation, the \emph{given clause algorithm} organises the set of clauses into two sets: the \Active\ clauses are those that have been active in inferences, whereas the \Passive\ clauses are those that have not. Typically, a further \Unprocessed\ set is required in order to manage the clauses produced during a single iteration of the loop. 
%
Realisations of the given clause algorithm generally differ in how they organise simplifications. There are two main approaches (both implemented by Vampire): the Otter loop uses both \Active\ and \Passive\ for simplifications, whereas the Discount loop only uses \Active.

The algorithm is centred around the \emph{clause selection} process. As previously mentioned, there are two main heuristics for this:
\begin{itemize}
	\item \emph{By Age (or First-in/First-out)} clause selection prefers the oldest clause (produced earlier in proof search), simulating a \emph{breadth-first} search of the clause space. In Vampire the age of a clause is the number of inferences performed to produce it (e.g. input clauses have age 0).
	\item \emph{By Weight (or symbol-counting)} clause selection prefers the youngest clause (with the fewest symbols). The intuition behind this approach is that the sought empty clause has zero symbols and lighter clauses are (in some sense) closer to this. Furthermore, lighter clauses are friendlier (less explosive) to proof search in various ways e.g. they are more general in terms of subsumption and tend to have fewer children.
\end{itemize}
It has been shown (see \sandm{}) that alternating these heuristics is beneficial and we consider this setting here. In Vampire this alternation is achieved by an age/weight ratio (AWR) implemented by a simple \emph{balancing} algorithm. The balance is initialised to 0 and used as follows: a positive balance means that a clause should be selected by age, whereas a negative balance means that a clause should be selected by weight; given a ratio of $a:w$ the balance is incremented by $a$ when selecting by age and decremented by $w$ when selecting by weight.

Figure~\ref{fig:discount} gives the Discount algorithm along with balance-based AWR clause selection. The lines relevant to clause selection are marked with \Mark.

@@

%%%%%%
\paragraph{Related Work.} We briefly review the clause selection approaches taken by other solvers. 
%
Otter \cite{} and Prover9 \cite{} CHECK BUT THEY PROBABLY  ALLOW BOTH.
%
E~\cite{E} allows the user to specify an arbitrary number of priority queues and a weighted round-robin scheme that determines how many clauses are picked from each queue. The default is to use a combination of age and weight selection although there is also a complex strategy developed by a genetic algorithm~\cite{DBLP:conf/gcai/SchaferS15}. 
%
SPASS~\cite{Spass} uses symbol-counting based clause selection. 
%
iProver~\cite{Iprover} follows in E in having a number of configurable queues but relies mainly on age and weight heuristics in those queues. 
%
The general idea in this paper of \emph{vary the ratio over time} can be applied to any ratio-based clause selection strategy and our specific results apply to those that take a ratio between age and weight.

%\newpage
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Optimising Age/Weight Ratios} \label{sec:awr:study}

@@

This experiment was repeated on the whole TPTP problem set, excluding problems Vampire does not currently support (e.g. higher-order problems). We ran Vampire for 1 second in default mode with the \texttt{discount} saturation algorithm\footnote{The default LRS~\cite{LRS} saturation algorithm can be non-deterministic so we avoid it for such experiments.} using a sensible set of AWR values (see Table \ref{tab:awr-improvement}) - these are the values used in Vampire's portfolio mode. These tend to favour weight-first over age-first as this has been experimentally shown to be preferable. We then filter out problems not solved by any of these or those solved trivially (e.g in preprocessing). The whole set yielded similar AWR data for 7,947 problems. 
%

The first result is that choosing a good AWR value for a problem is well-rewarded. Table~\ref{tab:awr-improvement} summarises the impact that choosing the best AWR can have. Compared to the default we can perform, on average, 1.26 times fewer activations, which is modest but (as Table \ref{tab:no-best-awr}) shows we also lose just under 10\% of problems by choosing the default. It is more relevant to note that there are cases where we can do \emph{much} better by selecting a different AWR value. So choosing a better AWR value \emph{can} go from no solution to a solution and \emph{can} do so faster (not necessarily). In the worst case we can do almost 4x worse on average by always choosing the worst AWR value.

The second result is that there is no ``best'' AWR across this full set of problems. Table \ref{tab:no-best-awr} shows, for each AWR value, the \% of problems solved, the number solved uniquely, the maximum and mean drop in performance. The drop in performance is the how many times more activations were required for a proof in the worst/average case compared to the best. No AWR value solves all problems, with the best being 1:5. 
A ratio of 1:4 produces an unusually well-behaved maximum performance drop.
\sandm~found that 1:5 had a similar property, but this might be explained by differences in prover and test environment. It is interesting to note that the extreme AWR values solve fewer problems overall but solve the most uniquely. This is typically in saturation-based proof search; approaches that do not perform well in general may perform well in specific cases where the general approach does not.

@@

\begin{table}[t]
	\caption{Per-AWR value results on 1 second runs over 7,947 TPTP problems.
%The mean relative performance drop (in terms of activations compared to an optimal value) choosing a single AWR over the problem set (approx. 8,000 problems solved in 1 second) causes.
%The percentage of problems in this set solved by the AWR setting is also shown.
%No setting is significantly ``better'' or ``worse'' than the others, but some have more variance in performance.
}
	\centering
	\begin{tabular}{l | c | c | c | c  c }
		AWR & \% Solved & Uniques & Maximum Drop & Mean Drop & (Standard Deviation)\\
\hline
8:1 & 85.25 & 16 & 150.67 & 1.37 & 1.98\\
5:1 & 86.10 & 1 & 122.22 & 1.33 & 1.64\\
4:1 & 86.93 & 1 & 101.44 & 1.32 & 1.42\\
3:1 & 87.63 & 2 & 105.00 & 1.29 & 1.41\\
2:1 & 88.62 & 3 & 112.67 & 1.27 & 1.45\\
3:2 & 89.83 & 2 & 119.89 & 1.27 & 1.51\\
5:4 & 89.98 & 4 & 125.00 & 1.26 & 1.55\\
1   & 90.56 & 4 & 133.56 & 1.26 & 1.63\\
2:3 & 91.20 & 9 & 147.67 & 1.28 & 1.79\\
2   & 91.68 & 0 & 162.67 & 1.31 & 1.97\\
3   & 91.81 & 5 & 190.56 & 1.37 & 2.30\\
4   & 91.85 & 3 & \textbf{17.41} & 1.38 & 0.67\\
5   & 92.00 & 2 & 133.67 & 1.44 & 1.70\\
6   & 91.57 & 1 & 106.44 & 1.47 & 1.46\\
7   & 91.49 & 1 & 104.89 & 1.49 & 1.44\\
8   & 91.09 & 2 & 101.33 & 1.53 & 1.45\\
10  & 90.52 & 1 & 101.78 & 1.60 & 1.53\\
12  & 90.00 & 0 & 101.67 & 1.65 & 1.62\\
14  & 89.29 & 4 & 103.00 & 1.70 & 1.75\\
16  & 89.42 & 5 & 101.33 & 1.74 & 1.76\\
20  & 88.61 & 3 & 100.89 & 1.82 & 1.94\\
24  & 88.26 & 2 & 101.33 & 1.89 & 2.08\\
28  & 87.57 & 2 & 99.22 & 1.96 & 2.24\\
32  & 87.01 & 1 & 100.00 & 1.99 & 2.36\\
40  & 86.23 & 4 & 98.78 & 2.09 & 2.64\\
50  & 84.93 & 1 & 98.78 & 2.17 & 2.88\\
64  & 84.17 & 2 & 101.22 & 2.28 & 3.19\\
128 & 81.34 & 3 & 107.44 & 2.57 & 4.16\\
1024 & 73.11 & 23 & 222.01 & 2.83 & 7.55
	\end{tabular}
	\label{tab:no-best-awr}
\end{table}

>>>>

\documentclass{llncs}
\usepackage{graphicx}
\usepackage[caption=false]{subfig}
\usepackage[dvipsnames]{xcolor}
\usepackage{amssymb}
\usepackage{amsmath}

@@

\begin{document}
\maketitle
\begin{abstract}
Modern saturation theorem provers are based on the given-clause algorithm, which iteratively selects new clauses to process. This clause selection has a large impact on the performance of proof search and has been the subject of much folklore.  
The standard approach is to alternate between selecting the \emph{oldest} clause and the \emph{lightest} clause with a fixed, but configurable \emph{age/weight ratio} (AWR).
We show that an optimal fixed value of this ratio can produce proofs significantly more quickly on a given problem, and further that varying AWR during proof search can improve upon a fixed ratio.
Based on these observations we develop several new modes for the Vampire prover  which vary AWR according to a ``shape'' during proof search.
The modes solve a number of new problems in the TPTP benchmark.
\end{abstract}

@@

Currently, the most successful theorem provers (such as Vampire~\cite{vampire}, E~\cite{E}, or SPASS~\cite{Spass}) for first-order logic are saturation-based, utilising the well-known \emph{given-clause algorithm}.
Simply, this algorithm \emph{saturates} a set of clauses by iteratively selecting a clause and performing all non-redundant inferences with it until all clauses have been selected or the empty clause (witnessing inconsistency) has been found.
Clearly, the order in which clauses are selected is key to the performance of the algorithm.
Over the past few decades a certain amount of folklore has built up around the best methods for clause selection and recent work by \sandm{} \cite{clause-selection-heuristics} systematically studied these.
Our work extends this study with new results and also introduces the concept of a \emph{variable} clause selection strategy (one that changes over time) which we instantiate with two simple patterns (or \emph{shapes}) that prove to be pragmatically useful.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Background}
\label{sec:background}
This section introduces the relevant background for the rest of the paper.

\paragraph{First-Order Logic and Weight.} Our setting is the standard first-order predicate logic with equality. A formal definition of this logic is not required for this paper but an important notion is that of the \emph{weight} of a clause. In first-order logic,  terms are built from function symbols and variables, literals are built from terms, and clauses are disjunctions of literals. The weight of a term is the number of function and variable symbols occurring in it and this notion is lifted to clauses.

\paragraph{Saturation-Based Proof Search.}
Saturation-based theorem provers \emph{saturate} a set of clauses $S$ with respect to an inference system $\mathbb{I}$: that is, computing a set of clauses $S'$ by applying rules in $\mathbb{I}$ to clauses in $S$ until no new clauses are generated. 
If the empty clause is generated then $S$ is unsatisfiable.
Calculi such as resolution and superposition have conditions that ensure \emph{completeness}, which means that a saturated set $S$ is satisfiable if it does not contain the empty clause as an element.
As first-order logic is only semi-decidable, it is not necessarily the case that $S$ has a finite saturation, and even if it does it may be unachievable in practice using the available resources. 
Therefore, much effort in saturation-based first-order theorem proving involves 
controlling proof search to make finding the empty clause more likely (within reasonable resource bounds). 
One important notion is that of \emph{redundancy}, being able to remove clauses from the search space that are not required. 
Another important notion are literal selections that place restrictions on the inferences that can be performed. 
Both notions come with additional requirements for completeness. Vampire often gives up these requirements for pragmatic reasons (incomplete strategies have been found to be more efficient than complete ones in certain cases) and in such cases the satisfiability of $S$ upon saturation is \emph{unknown}.

\paragraph{The Given Clause Algorithm and AWR Clause Selection.}
To achieve saturation, the \emph{given clause algorithm} organises the set of clauses into two sets: the \Active\ clauses are those that have been active in inferences, whereas the \Passive\ clauses are those that have not. Typically, a further \Unprocessed\ set is required in order to manage the clauses produced during a single iteration of the loop. 
Realisations of the given clause algorithm generally differ in how they organise simplifications. There are two main approaches (both implemented by Vampire, originally found in the eponymous theorem provers Otter~\cite{otter} and \textsc{Discount}~\cite{discount}): the Otter loop uses both \Active\ and \Passive\ for simplifications, whereas the Discount loop only uses \Active.

The algorithm is centred around the \emph{clause selection} process. As previously mentioned, there are two main heuristics for this:
\begin{itemize}
	\item \emph{By Age (or First-in/First-out)} clause selection prefers the oldest clause (produced earlier in proof search), simulating a \emph{breadth-first} search of the clause space. In Vampire the age of a clause is the number of inferences performed to produce it (e.g. input clauses have age 0).
	\item \emph{By Weight (or symbol-counting)} clause selection prefers the youngest clause (with the fewest symbols). The intuition behind this approach is that the sought empty clause has zero symbols and lighter clauses are (in some sense) closer to this. Furthermore, lighter clauses are more general in terms of subsumption and tend to have fewer children, making them less explosive in terms of proof search.
\end{itemize}
\sandm~show that alternating these heuristics is beneficial and we consider this setting here. In Vampire this alternation is achieved by an age/weight ratio (AWR) implemented by a simple \emph{balancing} algorithm. The balance is initialised to 0 and used as follows: a positive balance means that a clause should be selected by age, whereas a negative balance means that a clause should be selected by weight; given a ratio of $a:w$ the balance is incremented by $a$ when selecting by age and decremented by $w$ when selecting by weight.
Figure~\ref{fig:discount} gives the Discount algorithm along with balance-based AWR clause selection. The lines relevant to clause selection are marked with \Mark.

@@

%%%%%%
\paragraph{Related Work.}
We briefly review the clause selection approaches taken by other solvers. 
Otter 3.3~\cite{otter3} selects \emph{either} by age, by weight or manually.
Prover9~\cite{prover9} allows a configurable age/weight ratio.
E~\cite{E} allows the user to specify an arbitrary number of priority queues and a weighted round-robin scheme that determines how many clauses are picked from each queue.
The default is to use a combination of age and weight selection, although there is also a complex strategy developed by a genetic algorithm~\cite{DBLP:conf/gcai/SchaferS15}. 
SPASS~\cite{Spass} uses symbol-counting based clause selection. 
iProver~\cite{Iprover} follows in E in having a number of configurable queues but relies mainly on age and weight heuristics in those queues. 
The general idea in this paper of a \emph{varying age/weight ratio over time} is applicable to any ratio-based clause selection strategy, and our specific results apply to those that take a ratio between age and weight.

\section{Optimising Age/Weight Ratios}
\label{sec:awr:study}

@@

This experiment was repeated on the whole TPTP problem set, excluding problems Vampire does not currently support (e.g. higher-order problems). We ran Vampire for 1 second in default mode with the \texttt{discount} saturation algorithm\footnote{The default LRS~\cite{LRS} saturation algorithm can be non-deterministic so we avoid it for such experiments.} using a sensible set of AWR values (see Table \ref{tab:awr-improvement}) - these are the values used in Vampire's portfolio mode. These tend to favour weight-first over age-first as this has been experimentally shown to be preferable. We then filter out problems not solved by any of these or those solved trivially (e.g in preprocessing). The whole set yielded similar AWR data for 7,947 problems.

The first result is that choosing a good AWR value for a problem is well-rewarded.
Table~\ref{tab:awr-improvement} summarises the impact that choosing the best AWR can have.
Compared to the default we can perform, on average, 1.26 times fewer activations, which is modest but (as Table \ref{tab:no-best-awr}) shows we also lose just under 10\% of problems by choosing the default.
It is more relevant to note that there are cases where we can do \emph{much} better by selecting a different AWR value.
So choosing a better AWR value \emph{can} go from no solution to a solution and \emph{can} do so faster, but not necessarily.
In the worst case (choosing the pessimal AWR value) we can perform almost 4 times as many activations.

The second result is that there is no ``best'' AWR across this full set of problems. Table \ref{tab:no-best-awr} shows, for each AWR value, the \% of problems solved, the number solved uniquely, the maximum and mean drop in performance. The drop in performance is the how many times more activations were required for a proof in the worst/average case compared to the best. No AWR value solves all problems, with the best being 1:5. 
A ratio of 1:4 produces an unusually well-behaved maximum performance drop.
\sandm~found that 1:5 had a similar property, but this might be explained by differences in prover and test environment. It is interesting to note that the extreme AWR values solve fewer problems overall but solve the most uniquely. This is typical in saturation-based proof search: approaches that do not perform well in general may perform well in specific cases where the general approach does not.

@@

\begin{table}[t]
	\caption{
Per-AWR value results on 1 second runs over 7,947 TPTP problems.
	}
	\centering
	\begin{tabular}{l | c | c | c | c  c }
		AWR & \% Solved & Uniques & Maximum Drop & Mean Drop & (Standard Deviation)\\
\hline
8:1 & 85.25 & 16 & 150.67 & 1.37 & 1.98\\
5:1 & 86.10 & 1 & 122.22 & 1.33 & 1.64\\
4:1 & 86.93 & 1 & 101.44 & 1.32 & 1.42\\
3:1 & 87.63 & 2 & 105.00 & 1.29 & 1.41\\
2:1 & 88.62 & 3 & 112.67 & 1.27 & 1.45\\
3:2 & 89.83 & 2 & 119.89 & 1.27 & 1.51\\
5:4 & 89.98 & 4 & 125.00 & 1.26 & 1.55\\
1   & 90.56 & 4 & 133.56 & 1.26 & 1.63\\
2:3 & 91.20 & 9 & 147.67 & 1.28 & 1.79\\
2   & 91.68 & 0 & 162.67 & 1.31 & 1.97\\
3   & 91.81 & 5 & 190.56 & 1.37 & 2.30\\
4   & 91.85 & 3 & \textbf{17.41} & 1.38 & 0.67\\
5   & 92.00 & 2 & 133.67 & 1.44 & 1.70\\
6   & 91.57 & 1 & 106.44 & 1.47 & 1.46\\
7   & 91.49 & 1 & 104.89 & 1.49 & 1.44\\
8   & 91.09 & 2 & 101.33 & 1.53 & 1.45\\
10  & 90.52 & 1 & 101.78 & 1.60 & 1.53\\
12  & 90.00 & 0 & 101.67 & 1.65 & 1.62\\
14  & 89.29 & 4 & 103.00 & 1.70 & 1.75\\
16  & 89.42 & 5 & 101.33 & 1.74 & 1.76\\
20  & 88.61 & 3 & 100.89 & 1.82 & 1.94\\
24  & 88.26 & 2 & 101.33 & 1.89 & 2.08\\
28  & 87.57 & 2 & 99.22 & 1.96 & 2.24\\
32  & 87.01 & 1 & 100.00 & 1.99 & 2.36\\
40  & 86.23 & 4 & 98.78 & 2.09 & 2.64\\
50  & 84.93 & 1 & 98.78 & 2.17 & 2.88\\
64  & 84.17 & 2 & 101.22 & 2.28 & 3.19\\
128 & 81.34 & 3 & 107.44 & 2.57 & 4.16\\
1024 & 73.11 & 23 & 222.01 & 2.83 & 7.55
	\end{tabular}
	\label{tab:no-best-awr}
\end{table}
